Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-12-17 patchy-review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We thank the editor and the reviewers for their helpful comment. We
have substantially revised our paper in response to these comments,
and think that it is substantially improved as a consequence. Below we
have a detailed response point-by-point to the reviewers comments. We
first address the consensus points highlighted by the editor.

@@

\begin{quote}
  As can you see from the reviews, two general points were raised by all
  reviewers and I feel that thoroughly addressing both of these points poses a
  prerequisite for further consideration of the manuscript:

  (i) There needs to be some sort of numerical verification (e.g. with
  individual-based simulations) to test the main conclusions, check their
  robustness, and identify the parameter ranges for which they hold.
\end{quote}

We agree with this comment, to test the validity of our model we
ran an additional 21,873 simulations using our individual-based model of migration and selection,
and compared the results to the theory.
The main comparison occurs in \ref{fig:sim_times} and figures~\ref{fig:sim_probs};
we have also added in supplementary section \ref{apx:parameter_tables} 
visualizations of the simulations at various parameter values
that we think will help some readers in verifying the assumptions.

@@

\begin{quote}
  (ii) The presentation of the main result needs to be improved. In particular,
  all reviewers feel that the main result can be derived in a simpler and more
  intuitive way. I agree with them that at present the manuscript is overly
  technical and would benefit strongly from a more intuitive derivation of the
  results and discussion of the assumptions under which they hold. The reviewers
  provide detailed suggestions on how this can be achieved.
\end{quote}

We have rewritten the method section to simplify the presentation. 
Before writing the paper, we had initially explored a number of simpler ways of developing the
argument (along similar the lines to those suggested by the reviewers), 
and went around in circles for quite some time
before settling on the argument given in the first draft.
As suggested by the third reviewer, 
we have moved the most difficult part of the argument to the Materials \& Methods,
replacing it in the Results with a more \textit{ad hoc} argument
along the lines of that suggested by the reviewers.
We do think the more formal argument is more solid, however,
and therefore necessary to keep; 
as discussed below in our response to the reviewers, various seemingly reasonable arguments
give different answers for subtle reasons.

Other changes of note are:
\begin{itemize}

  \item Several reviewers pointed out that in approximating $\E[1-(1-p_e)^K] \approx \E[K] p_e$
    we were neglecting the case that $s_m \ll s_p$.
    This omission was sloppy on our part,
    which we have rectified; see the new equation~\eqref{eqn:migrate}.

  \item We have also improved our discussion of the limits to the theory,
    mostly in the (short) section \nameref{ss:assumptions}.
    The main point of note was that, indeed as Reviewer 3 pointed out,
    we were originally running simulations in regions where $s_m$ was too small for the theory to apply,
    resulting in $\migrate$ that was not monotonic with $s_m$.
    Happily, the natural restriction of $R \sqrt{s_m}/\sigma > 1$ coincides with the condition
    for $\migrate$ to decrease with $s_m$;
    we don't have good intuition for why the formula should increase in the region where it is inapplicable,
    but this seems beside the point.

@@

\end{itemize}

We hope you find the revised manuscript as improved as we do.

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

{\bf Peter Ralph, and\\
Graham Coop.}\\
\end{minipage}

\vskip 4em

@@

%%%%%%%%%%%%%%
\reviewersection

%%%%%%
\point{}{
    Am I right that the basic argument is the following:

    A migrant family in unfavorable territory will be typically of size $n \sim 1/s$
    ($s=s_m$), stick around for a time $\tau \sim 1/s$ and be spread out over $\ell \sim \sigma/\sqrt(s)$. 
    Given an equilibrium density $\rho*q(x)$, we need an arrival rate
    of migrant families of
    \[ k ~ \rho q(x) l/n/\tau \sim \rho q(x) l s^2 \]
}

\reply
We don't think the factor of $\ell$ should be there:
for example, suppose that $X$ is a Poisson process of ``centers'' of rate 1, 
and that around each center of $X$ we distribute $n$ points uniformly within a circle of radius $\ell$;
the expected number of points in a set of area 1 is $n$, regardless of $\ell$.
The right thing to do with the factor of $\tau$ is harder to intuit:
it is clear that $\rho q(x)$ should be $E[K]$ multiplied by the number of families present;
and it might be argued that the number of families present should be the influx of families
multiplied by the time they stick around for,
but given that the typical size of a family decreases once it arrives,
the correct way forward becomes non-intuitive, to us, at least.

@@

%%%%%%%%
\point{}{ 
    The probability of establishment of at least one member of family is then
    \[ P_{est} = 1 - \exp(-p n) \sim 1 - \exp(-p_e/s) \]
    where $p_e$ is the establishment probability of a single individual. The result
    for $\migrate$ will then depend on whether $p_e/s$ is large or small.
    For $p_e/s$ large, $\migrate = \rho q(x) s^2 \ell$,
    otherwise,
    $\migrate = \rho q(x) s \ell p_e$.

    The latter is the case discussed in the manuscript. The other is important too
    and should be discussed. 
}

\reply
As discussed above, we have incorporated this good point; see around equation~\eqref{eqn:migrate}.

@@

%%%%%%%%
\point{Fig 4:}{ Why not use natural units for the axis of these plots,
  $R\sqrt(s)/\sigma$, $\rho \mu A/s$.
  }

\reply
Good idea, but we've decided to cut this figure in the resubmission.

%%%%%%%%
\point{in the discussion on page 10 "probability of parallel adaptation.."}{
    It should be noted that many of the scenarios at high mutation influx are quite
    hypothetical. While the colonization of migrants might dominate over local
    adaptation, this argument assumes that the migrants are at equilibrium
    frequency. The transient to reach this equilibrium might be much longer than
    the time required for de novo local adaptation.
}

\reply
We don't know of an approximation to the time to reach migration-selection equilibrium
that isn't too hand-wavy,
unfortunately. 
However, we now point out in the section ``\nameref{ss:assumptions}'' that we're missing this factor,
and discuss there how the time lag to reach this equilbrium contributes to discrepancies between theory and simulations.

@@

%%%%%%%%%%%%%%  Reviewer 2
\reviewersection

\begin{quote}
Briefly, I think that the derivation of the main results (12), (21), and (25) can
be made much shorter and easier to understand (for both empiricists and the-
orists), and that there need to be simulations that quantitatively confirm these
results. Overly-detailed comments are below.
\end{quote}

\point{}{
    I think that the main result (12) can be derived much more simply, and
    also extended. By writing that the probability that a family that arrives
    at the patch is successful is approximately $ 2s_p \E [K] /\xi^2$, the authors have
    implicitly assumed that this quantity is $\ll 1$. But this implies that the
    fluctuations in the number of B individuals arriving at the patch can be
    ignored, since even a large fluctuation does not saturate the probability
    of success. This is valid for $s_m \gg s_p$. Given that the fluctuations can be
    ignored, (12) can be derived immediately from $q(y)$ -- simply apply Fick’s
    Law to $q(y)\rho$ to get that the flux of new B individuals is $\sigma^2 |q'(y)|\rho/2$, and
    then multiply by the success probability $2s_p/\xi^2$. (Actually, the result is
    slightly different from (12) -- see X below.)
}

@@

\reply
This argument implicitly treats each migrant individual as independent,
ignoring the correlations of arrival times induced by families.
This is the reason for explicitly defining families to be those who share an ancestor living outside of $r_0$
more recently than one living inside the patch;
many of the particles that Fick's law would pick up leaving $r_0$ would be already counted as part of a family.
However, we think our simplified argument in \nameref{ss:heuristics} gets the same spirit.

\point{}{
    However, I think we should also care about the opposite limit, $s_m \ll s_p$. In
    this case, the fluctuations do matter -- arrivals typically occur in clumps
    (families) large enough so that success is almost certain conditioned on
    arrival. Since on average each family consists of $\E[K]$ individuals, the
    family flux is just the individual flux divided by $\E[K]$. $\E[K] = \xi^2 /(2s_m)$,
    so the result is the same as above, except with $s_p$ replaced by an additional
    factor of $s_m$. (This could be made more formal, but I don’t think that’s
    what the authors are going for.) In the intermediate case $s_m \approx s_p$ which
    might be the most relevant for the mice, the two limiting regimes cross
    over smoothly. So for the first limit ($s_m \gg s_p$), all you need is $q$, and
    for the second, you only need to add basic facts about branching random
    walks (ie, limiting number distribution and localization at long times).
}

@@

%%%%%%%%
\point{"Length of the hitchhiking haplotype"}{
    I would focus more on the asymptotic expressions for $R \gg w  \gg \sigma/ \sqrt{2s_m}$, since that’s the regime
     where the analysis makes the most sense. Then the expressions become
     simpler and you can treat $d = 2$ as well. If you want, you could even
     lead with the intuitive explanation at the end of the section, and save the
     Bessel functions for an appendix.
}

 \reply
As discussed above, we've removed a big chunk of the section on haplotypes,
which should help make this section feel less heavy.
However, we don't quite see the reviewer's assertion that $R \gg w$ is important here,
and we don't see how to make the exposition much simpler with these assumptions.

%%%%%%%%
\point{}{ 
     I think (24) follows directly from Barton, Heredity 1979. (If you look at
     the unnumbered equation above Fig.\ 1 in that paper and plug in $x \sim a$,
     you find the time scale $r_{eff}$.) I think that the displayed equation should just
     be $r_{eff} \approx r\sigma/\sqrt{s}/a$. (I feel like the extra correction terms and geometric
      factor go against the spirit of the paper, and in any case are probably
      small compared to the inaccuracies introduced just by applying the model
      to any real system.) I think you can cut essentially everything before
      the simple explanation just above (24); if people want the mathematical
      details, they can read Barton. If I am misunderstanding things and there
    is a new result here, rather than just a good new exposition of an existing
    result, there should be a figure showing simulations confirming it.
}

\reply
This is a very interesting point,
and the basic tool of a diffusion approximation seems to be the same in both.
However, something is not matching up,
because \citet{barton1979geneflow} finds that the length of the region of differentiated genome 
decays with $1/\sqrt{T}$,
whereas with our approximation the length of the differentiated haplotype decays as $1/T$.
The difference may be that Barton considers LD and we consider a haplotype unbroken by recombination.
There are also other assumptions in Barton's paper that makes its applicability unclear.
On the balance, and considering the other reviewers' comments,
we've cut this section and discussed it a bit in the discussion. 
\llname{rr:haplotype_discussion}

%%%%%%%%
\point{In ``Applications'',}{the authors spend a lot of time introducing the mice
    and calculating the values for Fig. 6, but then they don’t show where the
    actual data lie on those plots. Is there enough information to do some
    kind of comparison? How close are the nearest samples to each other?
    Have no examples of alleles spreading by migration been found? If not,
    how much of a bound does this put on the parameter values? Is there any
    data on the genetic map length of the region of reduced diversity around
    the selected locus?
}

@@

\reply
At the moment it is only known that the same MC1R alleles do not
contribute to color polymorphism on a number of other patches. We know
from personal discussion with Michael Nachman that another color
polymorphism allele, in a different gene, has been mapped on another
patch. We are currently talking to Nachman about a broader-scale
effort to identify other causal alleles motivated in part by the work described
in the paper. However, this work is too preliminary to offer any thing
certain.

 There is no data about the genetic map in this species, but
as a rough rule of thumb 1cM is roughly 1Mb in a number of mammalian
species, we have added this guide to the physical scale of the shared haplotype in the figure caption.

%%%%%%%%
\point{}{ 
    A few sections or passages seem out of place:
    \begin{itemize}
        \item[A.]  I think ``Simulation methods'' should go in the ``Materials and methods'' section, or at least not right between ``Establishment of a locally
            adaptive allele due to mutational influx'' and ``Establishment of a lo-
            cally adaptive allele due to mutational influx''.

        \item[B.] The sentence on p7 beginning ``By our definition of fitness. . . ''.

        \item[C.]  The paragraph starting ``Also note that. . . '' on p11.
    \end{itemize}
}

@@

\reply
We moved \nameref{ss:simulations} to Materials and Methods,
modified the second sentence,
and rearranged the structure around the third paragraph.

%%%%%%%%
\point{Fig. 4:}{
    In the caption, I think it’s misleading to say that $\migrate$ is ``relatively insensitive to $s_m$''. 
    It’s exponentially sensitive! I don’t understand the blue numbers in the right-hand
    panel. It might be easier to interpret these if they were contour plots
    of the total time to adaptation, colored according to the value of (14),
    although I’m not sure. Also, they aren’t really ``phase diagrams''.
}

\reply
Good points; we meant that it was less sensitive to $s_m$ than if it didn't have the square root;
in this draft we've replaced this figure with a different one anyhow.

%%%%%%%%
\point{Top of p11 (just above ``Multiple patches''):}{should be $\sqrt{2s_m}/\sigma$.
}

@@

%%%%%%%%
\point{``the assumption of well-separated patches''}{ 
}

\reply
We have changed this. \llname{rr:separated_patches}

%%%%%%%%
\point{[figure 4, the contour plots]}{
}

\reply
Good point; we didn't notice that after correcting $\migrate$ the parameter regime on the right didn't make sense.
We have substituted figures \ref{fig:sim_times} and \ref{fig:sim_probs} for this one.

### 2014-12-17 patchy-selection-paper.tex
peter did final(?) pass through

   \newif\iflatexml\latexmlfalse %%%graham added

%We should cite Papers by Slatkin and Wiehe and this, on spatial hitchhiking http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3176130/
%%%%%
% Responses to reviews:
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\ifreviewresponses\else\renewcommand{\linelabel}[1]{}\fi
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review reponses file
\newcommand{\responsefile}{patchy-review-responses.tex}  % name of the review reponses file

@@

@@

%%%%%%%%%%%
\abstract{
Species often encounter, and adapt to, many patches of locally similar environmental conditions
across their range. 
Such adaptation can occur through convergent evolution as different alleles arise and spread in different patches, 
or through the spread of alleles by migration
acting to synchronize adaptation across the species.
The tension between the two reflects the degree of 
constraint imposed on evolution by the underlying genetic
architecture versus how effectively selection acts to inhibit the geographic spread of locally
adapted alleles.
This paper studies a model of the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a novel, locally adapted allele to appear in a
patch of habitat where it is favored through mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of distance between the patches?
How can we tell which has occurred, i.e.\ what population genetic
signal is left by the spread of migrant alleles and how long does this
signal persist for?
To answer these questions we decompose the migration--selection equilibrium
surrounding an already adapted patch
into families of migrant alleles,
in particular treating those rare families that reach new patches
as spatial branching processes.
This provides a way to understand the role of geographic separation between patches 
in promoting convergent adaptation and the genomic signals it leaves behind.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

\textbf{Author Summary:}
Often, a large species range will include patches where the species differs because it has adapted to locally differing conditions.
For instance, rock pocket mice are often found with a coat color that matches the rocks they live in, 
and the difference in coloration is known to be controlled genetically.
Sometimes, similar genetic changes have occurred independently in different patches,
suggesting that there were few accessible ways to evolve the locally adaptive form.
However, the genetic basis could also be shared if
rare migrants may carry the locally beneficial genotypes between nearby patches,
despite being at a disadvantage between the patches.
We use a mathematical model of random migration to determine how quickly adaptation is expected to occur through new mutation or migration from other patches,
and study in more detail what we would expect successful migrations between patches to look like.
The results are useful for determining whether similar adaptations in different locations are likely to have the same genetic basis or not,
and more generally in understanding how species adapt to patchy, heterogeneous landscapes.
}

@@

The convergent evolution of similar phenotypes 
in response to shared selection pressures is a testament to the power of selection 
to repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic,zhen2012parallel,martin2013repeated}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \citep{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ higher total mutational rate and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

%\citep{wright2013indirect}
One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{kruckeberg1951intraspecific,turner2010serpentine,macnair1991evolution,schat1996identical};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches.
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes
\citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which is on a number of black lava flows has much darker pelage than on intervening,
much lighter soil \citep{DiceBlossom1937}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible for the dark pigmentation found on different outcrops \citep{hoekstra2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism,dice1940ecologic}
and in lizards \citep{rosenblum2010molecular}.

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?
How easy will it be to detect adaptive alleles that are shared by
migration, i.e.\ over what genomic scale will a population genetic
signal be visible?

We work in a model of continuous geography,
using a variety of known results and new methods.
In the section \nameref{ss:patchymutation} we develop a simple approximation,
equation \eqref{eqn:mutrate}, for the rate at which patches become established by new mutations. 
The most important conceptual work of the paper occurs in the section \nameref{ss:patchymigration},
where we develop an understanding of the process by which alleles move from an existing patch 
to colonize a novel patch, culminating in equation \eqref{eqn:migrate} for the rate at which this happens.
We combine these two results in the section \nameref{ss:probparallel} 
to discuss to the probability of parallel adaptation, equation \eqref{eqn:parallel_prob}.
To understand the genomic signal left by adaptations shared by migration, 
in the section \nameref{ss:haplotype_length}, 
we approximate the time it will take for an allele to transit between patches,
(equation \eqref{eqn:mean_tau}),
and use this to approximate the length of haplotype that hitchhikes with it (equation \eqref{eqn:haplotype_length}). 
We then quantify how quickly this haplotype decays within patches (equation \eqref{eqn:rec_rate_within}) 
and use this to obtain properties about the equilibrium distribution of haplotype lengths 
(equation \eqref{eqn:within_haplotype_length}), in the section \nameref{ss:patch_haplotype}. 
Finally, in the section \nameref{ss:applications} we apply this work to understand the geographic scale of
convergent evolution in \textit{Chaetodipus intermedius}.

@@

Consider a population spread across a landscape to which it is generally well adapted,
but within which are patches of habitat to which individuals are (initially) poorly adapted.
(When we refer to ``patches'' it is to these pieces of poor habitat.)
Suppose it takes only a single mutational change to create an allele
($B$) that adapts an individual to the poor habitat type.
The required change could be as specific as a single base change, 
or as general as a knockout of a gene or one of a set of genes.
We assume that this mutational change occurs at a (low) rate of $\mu$ per chromosome per generation,
and that this initially rare new allele has fitness $1+s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches.
Outside of these patches we assume that the new allele has a relative fitness
$1-s_m$  when rare in the intervening areas, with $s_p$ and $s_m$ both positive.
(Here we define ``fitness'' to be the intrinsic growth rate of the type when rare.)
To simplify matters, we assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in these intervening regions.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

@@

Let $p(x)$ be the probability that a new mutant $B$ allele that arises at location $x$
relative to the center of the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approximation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to ignore migration, 
so that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This approximates the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is close to $p_e$ at some point and is symmetric about the edge of the patch. \linelabel{rr:p_estab}
We examine this more generally via exact calculation of $p(x)$ in the section \nameref{apx:establishment_sims}.

The rate at which mutations arise and colonize a patch of area $A$ is then
\begin{align} \label{eqn:mutrate} 
  \mutrate % = \rho \mu \int p(x) dx  
  \approx 2 s_p \rho A \mu / \xi^2,  
\end{align}
i.e.\ the product of mutational influx in the patch ($\rho A\mu$) and the probability of establishment ($2s_p/\xi^2$).
If this rate is low, then the time (in generations) until a mutation arises that
will become locally established within the patch is exponentially distributed with mean $1/\mutrate$.  
Assuming that once a mutation becomes established it quickly reaches its equilibrium frequency across the patch, 
the time scale on which new patches become colonized by the $B$ allele from new mutation is therefore approximately $1/\mutrate$.

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, 
study the fine-scale genealogy of alleles that transit between patches,
obtaining along the way other useful information about the genealogy
of $B$ alleles. \gc{Doing this we arrive at \eqref{eqn:migrate} for
  the rate at which patches adapt via migration. } %%Thought it good
                                %%to signpost again where we are
                                %%going.

>>>>

\newif\iflatexml\latexmlfalse %%%graham added; remove with latexml package

%%%%%
% Responses to reviews:
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\ifreviewresponses\else\renewcommand{\linelabel}[1]{}\fi
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review reponses file
\newcommand{\responsefile}{patchy-review-responses.tex}  % name of the review reponses file

@@

% make references to named sections look better
\newcommand{\secref}[1]{{\emph{\nameref{#1}}}}

@@

%%%%%%%%%%%
\abstract{
Species often encounter, and adapt to, many patches of locally similar environmental conditions across their range. 
Such adaptation can occur through convergent evolution if different alleles arise and spread in different patches, 
or through the spread of shared alleles by migration
acting to synchronize adaptation across the species.
The tension between the two reflects 
the degree of constraint imposed on evolution by the underlying genetic architecture 
versus how effectively selection and geographic isolation act to inhibit the geographic spread of locally adapted alleles.
This paper studies a model of the balance between these two routes to adaptation in continuous environments with patchy selection pressures.
We address the following questions: 
How long does it take for a novel, locally adapted allele to appear in a
patch of habitat where it is favored through mutation?
Or, through migration from another, already adapted patch?
Which is more likely to occur, as a function of distance between the patches?
How can we tell which has occurred, i.e.\ what population genetic
signal is left by the spread of migrant alleles?
To answer these questions we examine the family structure underlying migration--selection equilibrium
surrounding an already adapted patch,
in particular treating those rare families that reach new patches
as spatial branching processes.
This provides a way to understand the role of geographic separation between patches 
in promoting convergent adaptation and the genomic signals it leaves behind.
We illustrate these ideas using the convergent evolution of cryptic coloration in the rock pocket mouse, \textit{Chaetodipus intermedius}, as an empirical example.

\textbf{Author Summary:}
Often, a large species range will include patches where the species differs because it has adapted to locally differing conditions.
For instance, rock pocket mice are often found with a coat color that matches the rocks they live in,
these color differences are controlled genetically,
and mice that don't match the local rock color are more likely to be eaten by predators.
Sometimes, similar genetic changes have occurred independently in different patches,
suggesting that there were few accessible ways to evolve the locally adaptive form.
However, the genetic basis could also be shared if
migrants carry the locally beneficial genotypes between nearby patches,
despite being at a disadvantage between the patches.
We use a mathematical model of random migration to determine how quickly adaptation is expected to occur 
through new mutation and through migration from other patches,
and study in more detail what we would expect successful migrations between patches to look like.
The results are useful for determining whether similar adaptations in different locations are likely to have the same genetic basis or not,
and more generally in understanding how species adapt to patchy, heterogeneous landscapes.
}

@@

The convergent evolution of similar phenotypes 
in response to similar selection pressures is a testament to the power that selection 
has to sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes
\citep{stern2013genetic,zhen2012parallel,martin2013repeated}. 
Convergent adaptation also occurs within species, if different individuals
adapt to the same environment through different genetic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes \citep{Arendt:08}.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ higher total mutational rate and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
as this increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
which whether or not the copies derive from the same mutation,
leads to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

%\citep{wright2013indirect}
One elegant set of such examples is provided by the assortment of plant species 
that have repeatedly adapted to patches of soil with high concentrations of heavy metals
(e.g.\ serpentine outcrops and mine tailings) \citep{kruckeberg1951intraspecific,macnair1991evolution,schat1996identical,turner2010serpentine};
the alleles conferring heavy metal tolerance are thought to be locally restricted 
because they incur a cost off of these patches.
Similarly, across the American Southwest, a variety of species of animals have developed locally adaptive cryptic coloration
to particular substrates, e.g.\ dark rock outcrops or white sand dunes
\citep{benson1933concealing}.
One of the best-known examples is the rock pocket mouse (\textit{Chaetodipus intermedius}),
which on a number of black lava flows has much darker pelage than on intervening areas of lighter rock \citep{DiceBlossom1937}.
Strong predator-mediated selection appears to favour such crypsis \citep{kaufman1974adaptive},
and, perhaps as a result of this strong selection against migrants, 
at least two distinct genetic changes are responsible for the dark pigmentation found on different outcrops \citep{hoekstra2003different}. 
Similar situations have been demonstrated in other small rodent systems \citep{steiner2009genetic,kingsley2009melanism,dice1940ecologic}
and in lizards \citep{rosenblum2010molecular}.

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by inhabited regions where the alleles are deleterious.
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, 
i.e.\ convergently through new mutation,
and when is it likely that migration carries these alleles between patches?
How easy will it be to detect adaptive alleles that are shared by
migration, i.e.\ over what genomic scale will a population genetic
signal be visible?

We work in a model of continuous geography,
using a variety of known results and new methods.
In the section \secref{ss:patchymutation} we develop a simple approximation,
equation \eqref{eqn:mutrate}, for the rate at which patches become established by new mutations. 
The most important conceptual work of the paper occurs in the section \secref{ss:patchymigration},
where we develop an understanding of the process by which alleles move from an existing patch 
to colonize a novel patch, culminating in equation \eqref{eqn:migrate} for the rate at which this happens.
We combine these two results in the section \secref{ss:probparallel} 
to discuss to the probability of parallel adaptation, equation \eqref{eqn:parallel_prob}.
To understand the genomic signal left by adaptations shared by migration, 
in the section \secref{ss:haplotype_length}, 
we approximate the time it will take for an allele to transit between patches,
(equation \eqref{eqn:mean_tau}),
and use this to approximate the length of haplotype that hitchhikes with it (equation \eqref{eqn:haplotype_length}). 
Finally, in the section \secref{ss:applications} we apply this work to understand the geographic scale of
convergent evolution in \textit{Chaetodipus intermedius}.

@@

Consider a population spread across a landscape to which it is generally well adapted,
but within which are patches of habitat to which individuals are (initially) poorly adapted.
(When we refer to ``patches'' it is to these pieces of poor habitat.)
Suppose it takes only a single mutational change to create an allele
($B$) that adapts an individual to the poor habitat type.
The required change could be as specific as a single base change, 
or as general as a knockout of a gene or one of a set of genes.
We assume that this mutational change occurs at a (low) rate of $\mu$ per chromosome per generation,
and that this initially rare new allele has fitness advantage $s_p$ relative to the unmutated type ($b$) in these ``poor'' habitat patches.
Outside of these patches we assume that the new allele has a fitness disadvantage of 
$s_m$  when rare in the intervening areas, with $s_p$ and $s_m$ both positive.
(Here we define ``fitness'' to be the intrinsic growth rate of the type when rare.)
We assume that the disadvantage $s_m$ 
is sufficiently large that, on the relevant timescale,
the allele is very unlikely to fix locally in these intervening regions.
(The case where $s_m=0$ requires a different approach, which we do not treat here.)

@@

Let $p(x)$ be the probability that a new mutant $B$ allele that arises at location $x$
relative to the center of the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approximation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to ignore migration, 
so that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This approximates the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is close to $p_e$ at some point and is symmetric about the edge of the patch. \linelabel{rr:p_estab}
We examine this more generally via exact calculation of $p(x)$ in the section \secref{apx:establishment_sims}.

The rate at which mutations arise and colonize a patch of area $A$ is then
\begin{align} \label{eqn:mutrate} 
  \mutrate % = \rho \mu \int p(x) dx  
  \approx 2 s_p \rho A \mu / \xi^2,  
\end{align}
i.e.\ the product of mutational influx in the patch ($\rho A\mu$) and the probability of establishment ($p_e \approx 2s_p/\xi^2$).
If this rate is low, then the time (in generations) until a mutation arises that
will become locally established within the patch is exponentially distributed with mean $1/\mutrate$.  
Assuming that once a mutation becomes established it quickly reaches its equilibrium frequency across the patch, 
the time scale on which new patches become colonized by the $B$ allele from new mutation is therefore approximately $1/\mutrate$.

@@

Now suppose that there are two patches, each of area $A$ and separated by distance $R$. 
If the $B$ allele has arisen and become established in the first patch, but has not yet appeared in the second,
we would like to know the time scale on which copies of $B$ move between patches by migration
(supposing that no $B$ allele arises independently by mutation in the meantime).
To determine this, 
we study the fine-scale genealogy of alleles that transit between patches,
obtaining along the way other useful information about the genealogy of $B$ alleles. 
Doing this we arrive at \eqref{eqn:migrate} for the rate at which 
an allele established in one patch spreads to a neighboring one by migration.

Migration--selection balance ensures that 
there will be some number of $B$ alleles present in the regions where they are disadvantageous,
but only very few, and rarely, far away from the patch where $B$ is established.
Denote the expected frequency of allele $B$ at displacement $x$ relative to the patch by $q(x)$,
and assume that the shape of the patch is at least roughly circular.
Following \citet{haldane1948theory} or \citet{slatkin1973geneflow}, one can write down a differential equation to which $q(x)$ is the solution,
and show that $q(x)$ decays exponentially for large $|x|$,
with a square-root correction in two dimensions:
\begin{align} \label{eqn:eqfreq}
  q(x) \approx C \left( |x| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |x| \sqrt{2 s_m} / \sigma) \qquad \text{for large $|x|$},
\end{align}
where $d$ is the dimension ($d=1$ or $2$), 
and $C$ is a constant depending on the geographic shape of the populations and the selection coefficients.
In applications we fit $C$ to the data;
to get concrete numbers from theory we take $C=1$ if necessary.
